A causal analysis of the effect of abortion policies on maternal mortality

### 2022-06-08 PaperAnalysis.Rmd
Formatting

---
title: "Causal Analysis Paper"
author: "Helena Bakic, Anna-Lisa Wirth, Ángela Duarte Pardo"
date: "5/23/2022"
output: html_document
bibliography: references.bib
---

@@

# Background

Abortion law and the negative consequences of the lack of safe abortions are highly contentious policy issues that have been making headlines this year. Further, the [Sustainable Development Goal indicator 3.1](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/26) of the United Nations to reduce Maternal Death to less than 70 per 100000 life birth by 2030  is relying on access to safe and legal abortions. In the paper we are analysing the authors are testing whether and if so to what extent abortion law as a proxy for access to safe abortions causally effects maternal mortality rates in over 160 countries around the globe. Establishing a causal relation could strengthen the implementation of abortion liberalisation policy. This is especially needed in a time where several states already withdrew or are withdrawing access to safe abortions on many grounds such as [Poland](https://www.amnesty.org/en/latest/news/2021/10/poland-a-year-on-abortion-ruling-harms-women/) and the [USA](https://www.euronews.com/2022/05/03/us-supreme-court-moving-to-overturn-key-abortion-rights-law-says-leaked-report). While on the other hand there are developments like for instance in [France](https://www.hrw.org/news/2022/03/01/france-expands-abortion-access-two-key-moves) to expand abortion rights in 2022. So does abortion law flexibility cause maternal mortality? We will be taking a close critical look at a paper that tries to establish exactly that and also run our own data analysis to address this question.

# Description of the original paper

@@

#  Methodology

## Theorethical estimands

@@

The authors state that they aim to “estimate the association between the flexibility of abortion laws and maternal mortality at a country level [between 1985 and 2013]” [@Latt2019, pg. 2]. Furthermore, they estimated the model using linear regression, and the quantity they have estimated was the regression coefficient of MMR on flexibility of the law. Therefore, we we believe that the authors wanted to assess the average treatment effect (ATE), that is, the difference in the number of maternal deaths that would occur within a country if abortion law was flexible ($MMR_i$ = 1) versus inflexible ($MMR_i$ = 0):

@@

This estimand looks at counterfactual outcomes for all countries, regardless of the fact that the particular country would in reality receive treatment or not. Country $i$'s MMR when abortion laes are flexible is denoted as $MMR_i^{(L=1)}$ or shortly $MMR_i^{(1)}$, and as $MMR_i^{(L=0)}$ or $MMR_i^{(0)}$ when abortion laws are inflexible. Individual causal effect for the country $i$ is then the difference between mortality when abortion laws are flexible vs. when they are inflexible, or $ΔMMR_i = MMR_i^{(1)} - MMR_i^{(0)}$. Individual effects averaged across all of the countries in the world would then give the wanted causal quantity. This is equal to the expectation of the effect of flexible abortion laws on MMR of a randomly selected country.

@@

$$
\mathbb{E}[MMR^{1} \mid L = 1] = \mathbb{E}[MMR^{1} \mid L = 0]
$$
These assumptions are commonly violated in observational studies, as we will show in the next section, therefore it is to be expected that the three estimands (ATE, ATU, ATT) would be different in the current study.

## Identification

@@

However, there are several potential identification issues that could influence the possibility to estimate the effect of abortion law flexibility on MMR, namely; unmeasured confounding and (non-random) data missingness. These are discussed in detail below.

### Unmeasured confounding

What makes the estimation of ATE plausible is “uncoupling” or *d-separation* of assignment of units of analysis (in this case, countries) into treatment and control conditions. In randomized trials, this is achieved by randomly assigning treatment to unit analysis (Figure 2a). Randomization:

* Identifies the effect of Abortion law on MMR;
* Generated approximate balance in counfounders, even if a counfounder is unobserved;

which means that the causal effect of Law on MMR can be estimated. Conditioning on confounders is still (statictically) desirable, as it will increase the precision of the effect estimate.

@@

Given that unmeasured confounding is a crucial issue in effect identification, we have considered possible omitted variables and their role in the original model:

@@

* Religious views in society and level of secularism: The power of religious institutions and the views they hold in regard to the acceptability of abortion is likely to influence the level of maternal mortality and the flexibility of abortion law. If for instance a Christian Catholic institution has a lot of power in society and is very traditional this likely will influence the flexibility of the law but also the individual’s choice whether to get a clandestine abortion outside of the health system or whether she will even have the possibility to acquire a safe abortion in a health institution. However, this confounder would be extremely difficult to measure as there is no systematic recording of religious power and the views different institutions hold for all the studied countries.

@@

Lack of treating the missingness with female literacy

Comparing the rate of MMR and flexibility score for countries with recorded female primary completion vs countries without recorded female primary completion levels we could check whether there is a selection bias and then generalization of effects of this model would not be possible. If we suppose that female literacy is a collider and a confounder, we cannot condition on it without creating selection bias {@elwert2014endogenous, p.38}. As this done in the baseline paper, we would have to select strategies to correct this. Otherwise, it would stay d-connected, so statistically dependent. 
We see issues with potential interconnectedness and association of female literacy with abortion laws and access to abortion. If there is no school that can teach women how to read most likely there is no clinic that can perform a safe abortion, further women are unlikely to be informed about such services. While the issue of women’s information is taken up in the original paper and the lack of knowledge about abortion services even stays a problem for countries with high flexibility scores, low female primary completion here used as a better recorded proxy for female literacy is likely to cause access to abortion. So, it is a post-treatment confounder that potentially influences the outcome of MMR and abortion laws or the original variable of interest - access to safe abortions - it is a proxy for. The exact path between those variables should be analysed. The original paper suggested 2 models for this:

Model 1 included all the countries and no female literacy proxy:

```{r}
dag  <- dagify(MMrates ~ laws,
               MMrates ~ GDP,
               laws ~ GDP
               )
ggdag(dag) +  theme_dag_blank()
```

```{r}
dag  <- dagify(MMrates ~ laws,
               MMrates ~ GDP,
               MMrates ~ literacy,
               laws ~ GDP,
               laws~literacy
               )
ggdag(dag) +  theme_dag_blank()
```

However, it is also possible that female literacy has no effect on abortion law and actually only has an effect on the outcome, MMR. MMR is a common effect of female literacy and flexibility score (abortion laws). Then instead of being a causal effect to abortion law female primary completion would be an associative risk factor.{@10.2307/20485961, p.3}. In this case the only connection would be if we conditioned on MMR we would create a spurious relation however the effect of literacy and law on each other would be zero. Yet, we do not condition on the outcome hence no spurious relationship should be created.

```{r}
dag  <- dagify(MMrates ~ laws,
               MMrates ~ GDP,
               MMrates ~ literacy,
               laws ~ GDP
               )
ggdag(dag) +  theme_dag_blank()
```

However, it is difficult to disentangle the relationship as lower female literacy will also lower access to safe abortions regardless of the flexibility of the law as women are likely less informed about their health care options which in turn increases female mortality.

Testing for those different possible scenarios would be possible with sensitivity analysis or inverse probability weighting. We would need to control for the differences in the sets of considered countries (female lit = 1, recorded) and not considered countries (female literacy = 0, not-recorded) and see what other differences there are or trends in the groups other than female literacy being observed. While the authors drew two models they did not make this clear comparison in the paper. This additional analysis would help us to gain insight into whether recorded primary completion influences female mortality and to what extent. Further, it should be checked whether primary completion is a function of the state of human rights and GDP. The general state of human rights and equality could also influence abortion law and thus should be studied in order to make policy recommendations. However, a coded human rights and equality score that would be usable for causal analysis would need to be found. The IDEA democracy indices could be considered as a proxy for this (https://www.idea.int/data-tools/tools/global-state-democracy-indices).

@@

![Figure 1.: Suggested causal model for future research ](Figures/DAG_4.png)

```{r}
 testImplications <- function( covariance.matrix, sample.size ){
	library(ggm)
	tst <- function(i){ pcor.test( pcor(i,covariance.matrix), length(i)-2, sample.size )$pvalue }
tos <- function(i){ paste(i,collapse=" ") }
implications <- list(c("A","GDP","E","HRE"),
		c("A","GDP","E","L"),
		c("A","HRE","E","L"),
		c("E","L","HRE"),
		c("GDP","L","HRE"),
		c("HRE","MMR","A","E","GDP"),
		c("HRE","MMR","E","GDP","L"),
		c("L","MMR","A","E","GDP"),
		c("L","MMR","A","E","HRE"))
	data.frame( implication=unlist(lapply(implications,tos)),
		pvalue=unlist( lapply( implications, tst ) ) )

}
```

HRE: Human Rights and Equality
GDP : GDP
L : Abortion Law
A : Access to Abortion
E: Female Education
MMR : MMR

*Sensitivity Analysis*

In the sensitivity analysis we would check which part of the total effect can be explained by the abortion laws and which can be explained by female literacy. Additionally, we check how much the effects would need to change to cancel one another.For instance, the effect for flexibility score alone we can check when we control (block) the primary completion rate.
Sensitivity analysis has 2 essential parts to it:
1)describing the type of unobserved confounders
Unmeasured variables: religious institutions, contraceptive prevalence, fertility rate 
2) plausibility that problematic confounding might exist
	Based on 
	*research design, 
  *expert knowledge and 
  *data-generating process

Generally, sensitivity analysis is still underutilized in social science and policy research papers {@cinelli2020making, p. 40}. However, it is a very useful tool to test robustness of findings, i.e. how sensitive my effects and effect sizes are to other variables that influence the treatment or outcome studied. Therefore, we are proposing to add a sensitivity analysis to the research design and also apply it in the second part of our report. This method gives us two insights: 1) the robustness value: how robust the coefficient is to confounders influencing treatment and outcome and 2) the proportion of the outcome uniquely explained by the treatment and how strong a confounder would have to be to eliminate the effect of the treatment{@cinelli2020making, pp.40/41}.

*Missing data*
Hernan et al. are outlining in their paper why data is missing, to restrict the analysis on countries where data are complete would bias it. Countries with missing data on female literacy are still included {@10.2307/20485961, p. 5}. Female primary completion is only available for 50 percent of countries and female literacy only for 10 percent. This missingness of data introduces selection bias to this specific analysis the auhtors did in the original paper. As the likely selection bias in this study cannot be avoided due to the limitations on data availability it needs to be corrected if possible, in the analysis {@10.2307/20485961, p. 7} this could be achieved through generalization of inverse probability weighting. For this however we would need to categorize developing countries in one group where some have recorded female literacy which then gets multiplied by the number of countries not considered due to missing data. Inverse probability weighting would work as the score for female literacy is not zero for the countries where it is not recorded.  However, a lot of assumptions would need to be made to apply this method. The assumption that the filler-country would be close enough to the unobserved countries is by nature untestable and therefor it is a strong assumption.

## Empirical estimation

To estimate the effects of Abortion law flexibility on MMR, the authors of the paper fitted a fixed effects multiple regression model, in which they control for GDP, time, and countries, while trying to estimate the regression slope of flexible abortion law as a causal effect that is being studied:

@@

@@

G-separation is a method that enables the unbiased estimation of marginal causal effects by using counterfactuals. It  responds to two of the effect identification issues encountered in observational studies: the missingness of counterfactuals (because it is impossible to observe participants under both treatment and non-treatment) and heterogeneity of effects (variation of effects across individuals).[@snowden2011implementation]

G-separation calculation has two steps: 
1. Fitting a Q-model, that is, a traditional regression of the outcome on the treatment and  relevant covariates.  In our case, we used the traditional regression fitted in the original paper as Q model.

2. The Q-model is used to predict counterfactual outcomes for all the observations under both treatment settings (L=1 for flexible abortion laws and L=0 if not). More specifically, we plugged both treatment regimes in the the regression fit and obtained two predicted outcomes: $Y_{L=0}=Y_0$ for the unexposed and $Y_{L=1}=Y_1$ for the exposed. Both outcomes are then used to calculate the average $Y_1-Y_0$, which corresponds to the ATE.

With this intervention, the missingness problem is resolved because counterfactual outcomes for each observation are generated. G-separation thus provides a full dataset "free of confounding under causal assumptions."[@snowden2011implementation].

However, it is also worth noticing that G-separation also relies on correct model specification, as traditional regression models do, and on positivity and the existence and consistency of counterfactual outcomes (unlike regression). A clear advantage of the ATE estimator obtained by G-separation vis-à-vis regression is that it doesn't estimate the nuissance parameters and focuses only on the (causal) effect of interest.

Furthermore, G-separation allows for the calculation of both ATT and ATU by restricting the ATE estimation to the treated and the untreated respectively[@wang2017g].

# Analysis
Our empirical exploration had two main objectives: 1. to reproduce the methodology of the original paper with updated data on the same, 2. to compare the result thus obtained with some other methods that address the assumptions not met by the original paper.

*Updated datasets*
Although the datasets used in the original paper were hosted in public websites, the currently available version is not exactly the same. Since both the abortion policies and the maternal mortality sets are now hosted in interactive webpages, it was hard to access the aggregated and historical data from both abortion policies and maternal mortality rates between 1985-2013. Despite these restrictions, we collected data on all the relevant variables from 2000 to 2017. The maternal mortality data was accessed in the World Bank Data repository[@MMR], and the global abortion policies were found in the UN Department for Economic and Social Affairs archive [@WorldAbortionPolicies].

As in the original dataset, the Maternal mortality ratio (MMR) was measured as the number of maternal deaths per 100,000 live births and legal grounds for abortion where coded as binary variables ("1" for countries that legally allow for abortion on that ground; "0" for those that don't). We also calculated an aggregated and ordinal flexibility score for abortion by adding the legal grounds together.

@@

### Regression ATE
In the first part of our exploration, we followed the methodology of the original paper and used fixed effects linear regression models to estimate the relationship between abortion law flexibility and MMR. We controlled for GDP per capita, time in three-year intervals, and country-specific effects through a factor country-name variable. According to our results, flexible abortion laws reduce maternal deaths by 57 per 100.000 live births [ADD TABLE]. This result is close enough to the original result, which found that maternal deaths reduce by 45 per 100,000 live births (95% CI: -64, -26) with flexible abortion laws. 
We also conducted added "female primary education completion" as a control variable, and found very similar result: the effect of abortion law flexibility on maternal mortality appears to be a reduction in the mortality rate of 53 deaths per 100.000 live births.

*Heterogeinity of effects*

Following the approach of @Aronov2016, we have reproduced the weights that multiple regression analysis uses to caluclate the effects by 1. regressing the confounders and the fixed effects for countries in the analysis and the time period on the abortion law flexibility (measured as a score from 0 to 7) 2. squaring the weights.

As expected, our analysis showed that countries have different weights in the calculated effect. The nominal sample, the full population of the countries that have data for both the treatment and the outcome, is shown on Figure Xa and it includes almost all countries in the world. However, the effective sample, defined here as the countries that contribute over half (51%) of the weight used to construct the estimate of the abortion law flexibility on MMR, consists of only 14 countries (Figure Xb).

```{r, echo=FALSE,out.width="49%",out.height="49%",fig.show='hold', fig.align='center', fig.cap="Figure X. a) Nominal sample - all the countries in the analysis b) Effective sample - countries that contribute over 51% of the effect of abortion law flexibility on MMR"}

@@

As expected, countries with more data and higher variability in abortion law flexibility and MMR contribute more to the effect, as can be seen from the Figure X below.

```{r, echo=FALSE,out.width="49%",out.height="49%",fig.show='hold', fig.align='center', fig.cap="Figure X. a) Countries that contribute the most to the effect of abortion law flexibility on MMR vs. b) Countries that contribute the least"}

@@

*G-separation to find ATE, ATU, ATT*
After following the methodology outlined in @snowden2011implementation, we got an $ATE=-56,9$, which is roughly the same estimation of the effect of policy flexibility on MMR as the one resulting from the regression model. This is not surprising since both models estimated the effect as if it where identical across all strata of covariates because no covariate interactions were specified.

@@

The latter would mean that the potential impact of flexibility on abortion policies on MMR among those countries that have very stringent laws is almost non-existent, contradicting one of the central assumptions of the original paper. However, such a conclusion is not completely justified. On the contrary, this results comes with the need to revise the central assumptions under which the ATT and ATU can be calculated with G-separation. As explained by [@wang2017g], the computation of ATT and ATU also calls for consistency. This is, it requires assuming that the observed outcomes don't depend on the means by which the treatment was assigned. As @cole2009consistency, this assumption is not very hard to fullfil in observational studies (because there is no way to manipulate the treatment). In the specific case of our study, there are aspects of the way the laws are adopted (judicial review, constitutional mandate, etc.) that might affect the outcome.

*Unmeasured confounding*
We conducted a sensitivity analysis to get an idea of the strength of unmeasured counfounders in our model. These were not considered in the traditional regression estimates because to avoid bias, regression estimates require the assumption of no unobserved confounders.

We run the analysis for "female primary completion levels" and gdp  our regression models: 
1. The same we used for the g-separation analysis
2. A different regression model including primary completion levels as confounder.

@@

But how strong must they be to completely remove the effect in both models? The following countour plots show that, while gdp doesn't have a strong effect on MMR, in the second model a counfounder three times as strong as primary_completion would remove the causal effect of interest.

@@

# Conclusion 
The policy conclusion the original paper draws is justified. However, we could expand upon the methodology to get a more precise result and consider possible models as the limitation of models is not sufficiently justified in their paper. There are several ways to amend or extend the original research design. We presented suggestions in the theoretical part of our report like applying sensitivity analysis. We generated partial insights with our own data analysis. The causal relation between abortion law and maternal mortality ratios could be established. However, the proportion of the effect remains to be studied with an extended methodological toolbox.

# Contributions

@@

# References

>>>>

---
title: "A causal analysis of the effect of abortion policies on maternal mortality"
author: "Helena Bakic, Anna-Lisa Wirth, Ángela Duarte Pardo"
date: "5/23/2022"
output: html_document
bibliography: references.bib
---

@@

## Background

Abortion law and the negative consequences of the lack of safe abortions are highly contentious policy issues that have been making headlines this year. Further, the [Sustainable Development Goal indicator 3.1](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/26) of the United Nations to reduce Maternal Death to less than 70 per 100000 life birth by 2030 relies on access to safe and legal abortions. In the paper we are analysing, the authors are testing whether and, if so, to what extent abortion law as a proxy for access to safe abortions causally affects maternal mortality rates in over 160 countries around the globe. Establishing a causal relation could strengthen the implementation of abortion liberalisation policy. This is especially needed in a time where several states already withdrew or are withdrawing access to safe abortions on many grounds, such as [Poland](https://www.amnesty.org/en/latest/news/2021/10/poland-a-year-on-abortion-ruling-harms-women/) and the [USA](https://www.euronews.com/2022/05/03/us-supreme-court-moving-to-overturn-key-abortion-rights-law-says-leaked-report). While on the other hand there are developments like, for instance, in [France](https://www.hrw.org/news/2022/03/01/france-expands-abortion-access-two-key-moves) to expand abortion rights in 2022. So does abortion law flexibility cause maternal mortality? We will be taking a close critical look at a paper that tries to establish exactly that and also run our own data analysis to address this question.

## Description of the original paper

@@

##  Methodology

### Theorethical estimands

@@

The authors state that they aim to “estimate the association between the flexibility of abortion laws and maternal mortality at a country level [between 1985 and 2013]” [@Latt2019, pg. 2]. Furthermore, they estimated the model using linear regression, and the quantity they have estimated was the regression coefficient of MMR on flexibility of the law. Therefore, we believe that the authors wanted to assess the average treatment effect (ATE), that is, the difference in the number of maternal deaths that would occur within a country if abortion law was flexible ($MMR_i$ = 1) versus inflexible ($MMR_i$ = 0):

@@

This estimand looks at counterfactual outcomes for all countries, regardless of the fact that the particular country would in reality receive treatment or not. Country $i$'s MMR when abortion laws are flexible is denoted as $MMR_i^{(L=1)}$ or shortly $MMR_i^{(1)}$, and as $MMR_i^{(L=0)}$ or $MMR_i^{(0)}$ when abortion laws are inflexible. Individual causal effect for the country $i$ is then the difference between mortality when abortion laws are flexible vs. when they are inflexible, or $ΔMMR_i = MMR_i^{(1)} - MMR_i^{(0)}$. Individual effects averaged across all of the countries in the world would then give the wanted causal quantity. This is equal to the expectation of the effect of flexible abortion laws on MMR of a randomly selected country.

@@

$$
\mathbb{E}[MMR^{1} \mid L = 1] = \mathbb{E}[MMR^{1} \mid L = 0]
$$
These assumptions are commonly violated in observational studies, as we will show in the next section. Therefore it is to be expected that the three estimands (ATE, ATU, ATT) would be different in the current study.

### Identification

@@

However, there are several potential identification issues that could influence the possibility to estimate the effect of abortion law flexibility on MMR, namely: unmeasured confounding and (non-random) data missingness. These are discussed in detail below.

**Unmeasured confounding**

What makes the estimation of ATE plausible is the “uncoupling” or *d-separation* of assignment of units of analysis (in this case, countries) into treatment and control conditions. In randomized trials, this is achieved by randomly assigning treatment to unit analysis (Figure 2a). Randomization would allow the causal effect of Law on MMR to be estimated because it:

* Identifies the effect of Abortion law on MMR; and
* Generates approximate balance in counfounders, even if a counfounder is unobserved;

Conditioning on confounders is still (statistically) desirable, as it will increase the precision of the effect estimate.

@@

Given that *unmeasured confounding* is a crucial issue in effect identification, we have considered possible omitted variables and their role in the original model:

@@

* Religious views in society and level of secularism: The power of religious institutions and the views they hold in regard to the acceptability of abortion is likely to influence the level of maternal mortality and the flexibility of abortion law. If, for instance, a Christian Catholic institution has a lot of power in society and is very traditional, this will likely influence the flexibility of the law but also the individual’s choice whether to get a clandestine abortion outside of the health system or whether she will even have the possibility to acquire a safe abortion in a health institution. However, this confounder would be extremely difficult to measure as there is no systematic recording of religious power and the views different institutions hold for all the studied countries.

@@

**Missing data**

(Non-random) Data missingness is also a factor that can affect the possibility of accurately identifying causal effects.In the original paper, Hernan et al. are outlining why data is missing. Because restricting the analysis on countries where data are complete would bias it, the authors still include countries with missing data on female literacy [@HernanSelection, p. 5]. However, female primary completion is only available for 50 percent of countries and female literacy only for 10 percent. This missingness of data could introduce selection bias to the specific analysis the auhtors did in the original paper.

Comparing the rate of MMR and flexibility score for countries with recorded female primary completion vs countries without recorded female primary completion levels, we could check whether there is a selection bias and then generalization of effects of this model would not be possible. If we suppose that female literacy is a collider and a confounder, we cannot condition on it without creating selection bias [@elwert2014endogenous, p.38]. As this done in the baseline paper, we would have to select strategies to correct this. Otherwise, it would stay d-connected, so statistically dependent.

As the likely selection bias in this study cannot be avoided due to the limitations on data availability, it needs to be corrected if possible. In the analysis this could be achieved through generalization of inverse probability weighting [@HernanSelection, p. 7]. For this, however we would need to categorize developing countries in one group where some have recorded female literacy which then gets multiplied by the number of countries not considered due to missing data. Inverse probability weighting would work as the score for female literacy is not zero for the countries where it is not recorded.  However, a lot of assumptions would need to be made to apply this method. The assumption that the *filler-country* would be close enough to the *unobserved countries* is by nature untestable and therefore it is a strong assumption.

**Further association issues**

We see issues with potential interconnectedness and association of female literacy with abortion laws and access to abortion. If there is no school that can teach women how to read most likely there is no clinic that can perform a safe abortion, further women are unlikely to be informed about such services. While the issue of women’s information is taken up in the original paper and the lack of knowledge about abortion services even stays a problem for countries with high flexibility scores, low female primary completion here used as a better recorded proxy for female literacy is likely to cause access to abortion. So, it is a post-treatment confounder that potentially influences the outcome of MMR and abortion laws or the original variable of interest - access to safe abortions - it is a proxy for. The exact path between those variables should be analysed. The original paper suggested 2 models for this:

```{r, echo=FALSE,out.width="49%",out.height="49%",fig.show='hold', fig.align='center', fig.cap="Figure 3. a) Model 1 includes all the countries and no female literacy proxy , b) Model2: Model including female literacy as proxy"}

knitr::include_graphics(c("Figures/Dag_model1.png","Figures/Dag_model2.png"))
```

However, it is also possible that female literacy has no effect on abortion law and actually only has an effect on the outcome, MMR. MMR is a common effect of female literacy and flexibility score (abortion laws). Then instead of being a causal effect of abortion law, female primary completion would be an associative risk factor[@HernanSelection, p.3]. In this case the only connection would be if we conditioned on MMR we would create a spurious relation however the effect of literacy and law on each other would be zero. Yet, we do not condition on the outcome hence no spurious relationship should be created.

![Figure 4.:Female literacy's effect on Maternal mortality ratio](Figures/Dag_model3.png)

However, it is difficult to disentangle the relationship as lower female literacy will also lower access to safe abortions regardless of the flexibility of the law, as women are likely less informed about their health care options which in turn increases female mortality.

Testing for those different possible scenarios would be possible with sensitivity analysis or inverse probability weighting. We would need to control for the differences in the sets of considered countries (female lit = 1, recorded) and not considered countries (female literacy = 0, not-recorded) and see what other differences there are or trends in the groups other than female literacy being observed. While the authors drew two models they did not make this clear comparison in the paper. This additional analysis would help us to gain insight into whether recorded primary completion influences female mortality and to what extent. Further, it should be checked whether primary completion is a function of the state of human rights and GDP. The general state of human rights and equality could also influence abortion law and thus should be studied in order to make policy recommendations. However, a coded human rights and equality score that would be usable for causal analysis would need to be found. The [IDEA democracy indices](https://www.idea.int/data-tools/tools/global-state-democracy-indices) could be considered as a proxy for this.

@@

![Figure 5.: Suggested causal model for future research ](Figures/Dag_model4.png)

* Sensitivity Analysis

In the sensitivity analysis we would check which part of the total effect can be explained by the abortion laws and which can be explained by female literacy. Additionally, we check how much the effects would need to change to cancel one another.For instance, the effect for flexibility score alone we can check when we control (block) the primary completion rate.
Sensitivity analysis has 2 essential parts to it:
1) describing the type of unobserved confounders
Unmeasured variables: religious institutions, contraceptive prevalence, fertility rate 
2) plausibility that problematic confounding might exist
	Based on 
	*research design, 
  *expert knowledge and 
  *data-generating process

Generally, sensitivity analysis is still underutilized in social science and policy research papers [@cinelli2020making, p. 40]. However, it is a very useful tool to test robustness of findings, i.e. how sensitive my effects and effect sizes are to other variables that influence the treatment or outcome studied. Therefore, we are proposing to add a sensitivity analysis to the research design and also apply it in the second part of our report. This method gives us two insights: 1) the robustness value: how robust the coefficient is to confounders influencing treatment and outcome and 2) the proportion of the outcome uniquely explained by the treatment and how strong a confounder would have to be to eliminate the effect of the treatment[@cinelli2020making, pp.40/41].

### Empirical estimation

To estimate the effects of Abortion law flexibility on MMR, the authors of the paper fitted a fixed effects multiple regression model, in which they control for GDP, time, and countries, while trying to estimate the regression slope of flexible abortion law as the causal effect that is being studied:

@@

* Heterogeneous effects

@@

* Missingness of counterfactuals.

Even by controlling for the heterogeneous effects of treatment, there is still the lack of counterfactual problem that can become an issue when estimating causal effects in observational studies (because it is impossible to observe participants under both treatment and non-treatment). The G-separation method proposed by @snowden2011implementation addresses this issue by generating counterfactual outcomes for each observation and providing a full dataset "free of confounding under causal assumptions", which allows to isolate easily isolate causal effects (direct or heterogeneous)[@snowden2011implementation].

## Analysis
Our empirical exploration had two main objectives:

1. to reproduce the methodology of the original paper with updated data on the same covariates and outcome,

2. to compare the result thus obtained with some other methods that address heterogeneity of effects, missingness of confounders and unmeasured confounders.

**Updated datasets**

Although the datasets used in the original paper were hosted in public websites, the currently available version is not exactly the same. They were both hosted in interactive webpages, so we couldn't access the aggregated and historical data for both abortion policies and maternal mortality rates between 1985-2013. Despite these restrictions, we collected data on all the relevant variables from 2000 to 2017. The maternal mortality data was accessed in the World Bank Data repository [@MMR], and the global abortion policies were found in the UN Department for Economic and Social Affairs archive [@WorldAbortionPolicies].

As in the original dataset, the Maternal mortality ratio (MMR) was measured as the number of maternal deaths per 100,000 live births; legal grounds for abortion where coded as binary variables ("1" for countries that legally allow for abortion on that ground; "0" for those that don't), and the flexibility score for abortion policies was calculated by adding the legal grounds together.

@@

**Regression ATE**

In the first part of our exploration, we followed the methodology of the original paper and used fixed effects linear regression models to estimate the relationship between abortion law flexibility and MMR. We controlled for GDP per capita, time in three-year intervals, and country-specific effects through a factor country-name variable. According to our results, flexible abortion laws reduce maternal deaths by 57 per 100.000 live births. This result is close enough to the original result, which found that maternal deaths reduce by 45 per 100,000 live births (95% CI: -64, -26) with flexible abortion laws. 
We also conducted added "female primary education completion" as a control variable, and found very similar result: the effect of abortion law flexibility on maternal mortality appears to be a reduction in the mortality rate of 53 deaths per 100.000 live births.

**Treating the heterogeinity of effects**

Following the approach of @Aronov2016, we have reproduced the weights that multiple regression analysis use to calculate the effects by:

1. regressing the confounders and the fixed effects for countries in the analysis (without the treatment) and the time period on the abortion law flexibility score (treatment measured as a score from 0 to 7) 
2. squaring the weights.

As expected, our analysis showed that countries have different weights in the calculated effect. The nominal sample, the full population of the countries that have data for both the treatment and the outcome, is shown on Figure 6a and it includes almost all countries in the world. However, the effective sample, defined here as the countries that contribute over half (51%) of the weight used to construct the estimate of the abortion law flexibility on MMR, consists of only 14 countries (Figure 6b).

```{r, echo=FALSE,out.width="49%",out.height="49%",fig.show='hold', fig.align='center', fig.cap="Figure 6. a) Nominal sample - all the countries in the analysis b) Effective sample - countries that contribute over 51% of the effect of abortion law flexibility on MMR"}

@@

As expected, countries with more data and higher variability in abortion law flexibility and MMR contribute more to the effect, as can be seen from  *Figure 7* below.

```{r, echo=FALSE,out.width="49%",out.height="49%",fig.show='hold', fig.align='center', fig.cap="Figure 7. a) Countries that contribute the most to the effect of abortion law flexibility on MMR vs. b) Countries that contribute the least"}

@@

**Treating confounders' missingness through G-separation**

G-separation, as proposed by @snowden2011implementation, starts with fitting traditional regression of the outcome on the treatment and relevant covariates. It then uses it to predict counterfactual outcomes for all the observations under both treatment settings (L=1 for flexible abortion laws and L=0 if not). More specifically, by plugging both treatment regimes in the regression fit, two predicted outcomes are obtained: $MMR^{(L=0)}=MMR^{0}$ for the unexposed and $MMR^{(L=1)}=MMR^{1}$ for the exposed. Both outcomes are then used to calculate the average $MMR^{1}-MMR^{0}$, which corresponds to the ATE. Furthermore, G-separation allows for the calculation of both ATT and ATU by restricting the ATE estimation to the treated and the untreated respectively[@wang2017g].

After following this metodology, we got an $ATE=-56,9$, which is roughly the same as the result of the regression model. This is not surprising since both models estimated the effect as if it where identical across all strata of covariates because no covariate interactions were specified, which would account for heterogeneous effects.

@@

The latter would mean that the potential impact of flexibility on abortion policies on MMR among those countries that have very stringent laws is almost non-existent, contradicting one of the central assumptions of the original paper. However, such a conclusion is not completely justified. On the contrary, this results comes with the need to revise the central assumptions under which the ATT and ATU can be calculated with G-separation. As explained by [@wang2017g], the computation of ATT and ATU also calls for consistency. This is, it requires assuming that the observed outcomes don't depend on the means by which the treatment was assigned. As @cole2009consistency, this assumption is very hard to fullfil in observational studies (because there is no way to manipulate the treatment). In the specific case of our study, there are aspects of the way the laws are adopted (judicial review, constitutional mandate, etc.) that might affect the outcome.

**Using sensitivity analysis as an approach to unobserved confounding**

Finally, we conducted a sensitivity analysis to estimate the fragility of our the effect of interest against unobserved counfounders in our model[@cinelli2020making]. These potential confounders are not usually considered in  traditional regression estimates to avoid bias, but they might substantively change our conclusions.

We then run the sensitivity analysis proposed by @cinelli2020making using "female primary completion levels" and gdp as benchmark covariates for two models: 
1. The one used by the original paper (and used by us for the g-separation analysis).
2. A different regression model including primary completion levels as confounder.

@@

But how strong must they be to completely remove the effect in both models? The following countour plots show that, while gdp doesn't have a strong effect on MMR, in the second model a counfounder three times as strong as primary completion would remove the causal effect of interest. This means that even if the effect of primary completion levels appears to be small, other strong confounders as the ones listed above might easily reduce the effect of interest of our study.

@@

## Conclusion 
The policy conclusion the original paper draws is justified. However, we could expand upon the methodology to get a more precise result and consider possible models as the limitation of the approach is not sufficiently justified in their paper. There are several ways to amend or extend the original research design. We presented suggestions in the theoretical part of our report like applying sensitivity analysis and accounting for the heterogeneity of effects. We generated partial insights with our own data analysis. The causal relation between abortion law and maternal mortality ratios could be established. However, the proportion of the effect remains to be studied with an extended methodological toolbox.

## Contributions

@@

## References
